Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-10-05 figures.Rmd
plot all brain tracings

load("glm_model.Rdata")
load("results.Rdata")
load("roc_curves.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")

@@

  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
    as_tibble() %>% 
    dplyr::select(-1) %>% 
    mutate(Status = ifelse(Status == "HV", "HC", Status)) %>% 
    mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))

@@

# extract variables of interest
voi <- df_results %>% colnames() %>% .[4:9]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% 
  dplyr::filter(variable %in% voi) %>% 
  mutate(variable = factor(variable, levels = c("avg_Hb_o2sat", "avg_Hb_conc",
                                                "Hb_o2sat_overall_alpha", "Hb_conc_overall_alpha",
                                                "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))

@@

# resample
set.seed(234)
hbox_boot <- bootstraps(df_hbox_model, strata = Status, times = 1000)

@@

table2b <- hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
  mutate(p.value = scientific(p.value, digits = 3),
         std.error = round(std.error, 2),
         statistic = round(statistic, 2)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>% 
  as.data.frame()

@@

### Figure 6: Hemoglobin concentration trends in extravascular edema vs vascular engorgement

```{r fig6, fig.width = 12}

>>>>

load("glm_model.Rdata")
load("results.Rdata")
load("roc_curves.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")
load("all_dfa_results.Rdata")

@@

  df_hbox <- read.csv("Data/master_datatable.csv") %>% 
    as_tibble() %>% 
    dplyr::select(-1) %>% 
    mutate(Status = replace(Status, Status == "HV", "HC")) %>% 
    mutate(Status = factor(Status, levels = c("HC", "UM", "CM")))

@@

# extract variables of interest
voi <- df_results %>% colnames() %>% .[3:8]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% 
  dplyr::filter(variable %in% voi) %>% 
  mutate(variable = factor(variable, levels = c("avg_Hb_o2sat", "avg_Hb_conc",
                                                "Hb_o2sat_overall_alpha", "Hb_conc_overall_alpha",
                                                "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

@@

  df_hbox_model <- df_hbox_log_reg %>% 
    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
    mutate(`1.10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1) %>% 
    mutate(`1.10_Hb_conc_alpha_z` = (Hb_conc_second_alpha - mean(Hb_conc_second_alpha)) / sd(Hb_conc_second_alpha), 
           Hb_conc_z = (avg_Hb_conc - mean(avg_Hb_conc)) / sd(avg_Hb_conc),
           Hematocrit_z = (Hematocrit - mean(Hematocrit)) / sd(Hematocrit)) %>% 
    dplyr::select(-c(Hb_conc_second_alpha, `1.10_Hb_conc_alpha`, avg_Hb_conc, Hematocrit)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))

# resample
set.seed(234)
hbox_boot <- bootstraps(df_hbox_model, strata = Status)#, times = 1000)

@@

table2b <- hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
  mutate(p.value = scientific(p.value, digits = 3),
         std.error = round(std.error, 2),
         statistic = round(statistic, 2)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "t-statistic" = "statistic", "p-value" = "p.value") %>% 
  as.data.frame()

@@

### Figure 5: Cerebral malaria patient follow-up visits

```{r fig5_data}

# reorganize for plotting
df_results_all_plot <- df_brain_dfa_results %>% 
  dplyr::filter(!is.na(avg_Hb_conc_fu)) %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))

# descriptive statistics

  # averages

    # THC
    df_thc_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_conc", "avg_Hb_conc_fu"))

    thc_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_compare) %>% .$p.value

    # O2 sat
    df_o2sat_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_o2sat", "avg_Hb_o2sat_fu"))

    o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value

  # alphas

    # THC
    df_thc_alpha_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))

    thc_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_alpha_compare) %>% .$p.value

    # O2 sat
    df_o2sat_alpha_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("Hb_o2sat_overall_alpha", "Hb_o2sat_fu_alpha"))

    o2sat_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_alpha_compare) %>% .$p.value

#### A: Cerebral hemoglobin oxygen saturation & hemolgobin concentration
```{r fig5a}

# create color vector
my_col_fu <- c("initial visit" = "#B80F0A", "follow up" = "black")

# plot
df_results_all_plot %>%

  dplyr::filter(grepl("avg_Hb", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +

  facet_wrap(~measurement, scales = "free_y") +
  labs(y = "", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

```
No difference from initial visit to follow up - Hb_oxy pval: 0.90; Hb_tot pval: 0.32

#### B: Cerebral hemoglobin oxygen saturation & hemolgobin concentration alphas
```{r fig5b}

library(scales)

df_results_all_plot %>%

  dplyr::filter(!grepl("avg|second", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  geom_segment(aes(x = 1, y = 0.93, xend = 2, yend = 0.93), size = 0.3) +
  geom_segment(aes(x = 1, y = 0.93, xend = 1, yend = 0.89), size = 0.3) +
  geom_segment(aes(x = 2, y = 0.93, xend = 2, yend = 0.89), size = 0.3) +
  geom_text(x = 1.5, y = 0.96, aes(label = paste0("p = ", pvals))) +

  facet_wrap(~measurement) +

  ylim(0.45, 1.1) +
  labs(y = "alpha", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

### Figure 6: Muscle hemoglobin concentration and oxygen saturation data from initial visit and follow-up

#### A. Muscle hemoglobin concentration and oxygen saturation

```{r fig6A}

# select from the long df
df_muscle_long <- df_muscle_dfa_results %>% 
  dplyr::select(c(Status, avg_Hb_conc_muscle, avg_Hb_o2sat_muscle)) %>% 
  pivot_longer(2:3, names_to = "variable") %>% 
  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
         variable = factor(variable, levels = c("avg_Hb_o2sat_muscle", "avg_Hb_conc_muscle")))

# plot
df_muscle_long %>% dplyr::filter(Status != "?") %>%

```{r fig6a.pvals, eval = FALSE}

# hemoglobin concentration
kruskal.test(avg_Hb_conc_muscle ~ Status, data = df_muscle_dfa_results)

# hemoglobin oxygen saturation
kruskal.test(avg_Hb_o2sat_muscle ~ Status, data = df_muscle_dfa_results)

```
No significant differences: Hb_oxy pval = 0.11 by KW; Hb_tot pval = 0.051 by KW

#### B. Muscle DFA results
```{r fig6B}

# select from the long df
df_muscle_dfa_long <- df_muscle_dfa_results %>% 
  dplyr::select(c(Status, muscle_Hb_conc_alpha, muscle_Hb_o2sat_alpha)) %>% 
  pivot_longer(2:3, names_to = "variable") %>% 
  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
         variable = factor(variable, levels = c("muscle_Hb_o2sat_alpha", "muscle_Hb_conc_alpha")))

# plot
df_muscle_dfa_long %>% dplyr::filter(Status != "?") %>%

  facet_wrap(~variable) +
  ylim(0, 1.5) +

```{r fig6b.pvals, eval = FALSE}

# hemoglobin concentration alpha
kruskal.test(muscle_Hb_conc_alpha ~ Status, data = df_muscle_dfa_results)

# hemoglobin oxygen saturation alpha
kruskal.test(muscle_Hb_o2sat_alpha ~ Status, data = df_muscle_dfa_results)

```
No significant differences: Hb_oxy pval = 0.08 by KW; Hb_tot pval = 0.15 by KW

#### C. CM follow-up muscle results
```{r fig6c_data}

# reorganize for plotting
df_muscle_fu_plot <- df_muscle_dfa_results %>% 
  dplyr::filter(!is.na(avg_Hb_o2sat_muscle_fu)) %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))

# descriptive statistics

  # averages

    # THC
    df_thc_muscle_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_conc_muscle", "avg_Hb_conc_muscle_fu"))

    thc_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_compare) %>% .$p.value

    # O2 sat
    df_o2sat_muscle_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_o2sat_muscle", "avg_Hb_o2sat_muscle_fu"))

    o2sat_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_compare) %>% .$p.value

  # alphas

    # THC
    df_thc_muscle_alpha_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("muscle_Hb_conc_alpha", "muscle_Hb_conc_fu_alpha"))

    thc_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_alpha_compare) %>% .$p.value

    # O2 sat
    df_o2sat_muscle_alpha_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("muscle_Hb_o2sat_alpha", "muscle_Hb_o2sat_fu_alpha"))

    o2sat_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_alpha_compare) %>% .$p.value

Muscle hemoglobin concentration and hemoglobin oxygen saturation
```{r fig6c_avgs}

df_muscle_fu_plot %>%

  dplyr::filter(grepl("avg_Hb", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_muscle_fu_pval, 3), round(o2sat_muscle_fu_pval, 3))) %>% 
  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +

  facet_wrap(~measurement, scales = "free_y") +
  labs(y = "", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

```
No difference in hemoglobin concentration from initial visit to follow up - Hb_tot pval: 0.72. However, there was a significant drop in muscle hemoglobin oxygen saturation at the follow-up compared to initial visit (p = 0.044).

Muscle DFA results
```{r fig6c_dfa}

df_muscle_fu_plot %>%

  dplyr::filter(!grepl("avg_Hb", variable) & !grepl("second", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_muscle_fu_pval, 3), round(o2sat_muscle_fu_pval, 3))) %>% 
  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +

  facet_wrap(~measurement) +
  ylim(0.45, 1.05) +
  labs(y = "", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

```
No difference from initial visit to follow up - Hb_oxy pval: 0.17; Hb_tot pval: 0.10.

### Figure 7: Hemoglobin concentration trends in extravascular edema vs vascular engorgement

```{r fig7, fig.width = 12, eval = FALSE }
